About p-adic stability in linear algebra

### 2014-12-12 examples/stability.tex
More work on vector spaces

\noindent
{\bf Organization of the paper.}
The section \ref{sec:theory} introduces the theory of ultrametric 
precision developped in \cite{caruso-roe-vaccon:14a} and completes it 
on a technical point. In \S \ref{sec:matrices}

The analysis of standard algorithms are treated in 
the next sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
successively devoted to the case of numbers, matrices, univariate 
polynomials, multivariate polynomials and finally vector spaces.

@@

In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be made explicit in terms of the
growing function $\Lambda(f)$ of $f$ defined by:
$$\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)$$
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement and seek here to state the particular case of
polynomial functions. A function $f : E \to F$ is said \emph{polynomial} 
if it is coordinate-wise given by multivariate polynomials in any 
(equivalently all) system of coordinates associated to orthonormal 
basis.

\begin{prop}
\label{prop:precision2}
We keep the notations of Proposition \ref{prop:precision} and assume 
in addition that $f$ is polynomial. Let $C$ be a positive real
number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
\rho^{-1}$.
\end{prop}

@@

We actually generalize a bit the setting above and consider the 
differential equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
normed vector spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a $K$-Banach space $G$. Consequently, $g$
goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
all (continuous) linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$ and second, in order to 
apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
needs to be surjective and therefore \emph{a fortiori} nonzero.

@@

Actually, we do recommend taking $h$ as the identity function 
\emph{but} changing the norm on the codomain of $h$ at the same time.
More precisely, if $F$ --- which is the $K$-Banach in which $f$ takes
its values --- is equipped with the norm $\Vert \cdot \Vert_F$, we 
recommand endowing it with the second norm defined by
$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
where $\lambda$ is a positive real number that we shall choose later 
and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
acting as the identity on the underlying vector spaces. Doing this, the 
function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
choose $\Lambda_h = \Lambda(h)$.

@@

\subsection{Multiplication}

To begin with, we want to study the behaviour of the precision when 
performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
positive integers with $s \geq \min(r,t)$ and assume that we want to 
multiply a matrix $A \in M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. 
This operation is of course modeled by the polynomial function:
$$\begin{array}{rcl}
\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
M_{r,t}(K) \smallskip \\
(A,B) & \mapsto & AB.
\end{array}$$
According to Proposition \ref{prop:precision}, the behaviour of the precision when 
computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
that this differential is the linear mapping that takes a couple 
$(dA,dB)$ to $A \cdot dB + dA \cdot B$.

To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
Nonetheless, it may happen that the above inclusion is strict, meaning 
that we are \emph{gaining} precision in those cases. In order to compute 
exactly $\mathcal P'_{a,b,c}(A,B)(\mathcal L_0)$, we introduce the Smith 
factorization of $A$ and $B$:
$$A = U_A \cdot \Delta_A \cdot V_A
\text{ and } B = U_B \cdot \Delta_B \cdot V_B$$
and denote by $a_1, \ldots, a_{\min(r,s)}$ (resp. $b_1, \ldots, 
b_{\min(s,t)}$) the valuation of the diagonal entries of $\Delta_A$
(resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}(\O_K)$ as 
the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.

@@

Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
perspective of precision. The second statement roughly means that when 
we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
note that, on the other side, the valuation of the entries of $AB$ may 
increase, meaning that we are also loosing some significant digits if 
we are reasoning in relative precision.} (under the assumption of 
Proposition \ref{prop:precision2}). However these digits are not --- 
even individually --- attached to some particular entry of $AB$ but are 
``spread out'' among all its entries. Thus, a usual coordinate-wise 
precision data does not see this gain of precision in general.
There is a more pedestrian way to visualize this phenomenon. Remark that
the product $AB$ can be rewritten:
$$AB = U_A \cdot P \cdot V_B
\quad \text{with} \quad 
P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$$
Tracking precision in the usual way, we see that the $(i,j)$-th entry
of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
gain of precision is visible here. However it is then ``spread out'' 
when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
the usual coordinate-wise precision model. The lattice precision model 
then appears as a way to keep track of this ``diffused'' gain of
precision.

@@

\noindent
Figure \ref{fig:mulmatrix} compares the number of significant digits in 
\emph{relative} precision we are loosing on the output of Algorithm 1 
when we are using, on the one hand, a standard component-wise track of 
precision and, on the other hand, a lattice-based method to handle 
precision.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{2}{|c|}{Average loss of precision} \\
\cline{3-4}
& & Comp.-wise method & Lattice method \\
\hline 
$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
$2$ & $1000$ & $157.8$ & $7.9$ \\
\hline
$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
$3$ & $1000$ & $122.5$ & $7.0$ \\
\hline
\end{tabular}

@@

\noindent
{\bf Direct images.}
Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
interested here in the function $\DI$ (for ``direct image'') defined 
over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
priori} vary, the map $\DI$ does not take its values in well-defined 
grassmannian. To tackle this problem, we stratify $\mathcal M$ as
follows: for each integer $r \in [0,d]$, we introduce the subset
$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
locally closed in $\mathcal M$ are therefore are submanifolds. 
Moreover, $\DI$ induces by restriction respectable differentiable 
functions between $K$-manifolds:
$$\DI_r : \mathcal M_r \to \Grass(F,r).$$
We would like to differentiate $\DI_r$ around some point $(f,V) \in 
\mathcal M_r$. To do so, we use the first description of the 
grassmannians we gave above: we shall see points in $\Grass(E,d)$ 
(resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
\hookrightarrow F$) modulo the action of $\GL(V)$ (resp. $\GL(W)$).
The point $V \in \Grass(E,d)$ is then represented by the canonical 
inclusion $v : V \to E$ whereas a representative $w$ of $W$ satisfies
$w \circ \varphi = f \circ v$
where $\varphi : V \to W$ is the linear mapping induced by $f$. The 
previous relation still holds if $(f,v)$ is replaced by a couple $(f', 
v') \in \mathcal M_r$ sufficiently close to $(f,v)$.
Differentiating it and passing to the quotient we find, first, that the 
tangent space of $\mathcal M_r$ at $(f,v)$ consists of couples $(df, dv) 
\in \Hom(E,F) \times \Hom(V,E/V)$ such that
$$d\tilde w = \big((df \circ v + f \circ dv) \text{ mod } W\big)
\in \Hom(V, F/W)$$
factors through $\varphi$ (\emph{i.e.} vanishes on $\ker \varphi = V 
\cap \ker f$) and, second, that the differential of $\DI_r$ at $(f,V)$ 
is the linear mapping sending $(df,dv)$ as above to the unique element 
$dw \in \Hom(W,F/W)$ such that $dw \circ \varphi = d \tilde w$.

@@

\noindent
{\bf Inverse images.}
We now consider the mapping $\II$ (for ``inverse image'') sending a 
couple $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
$f^{-1}(W)$. As before, this map does not takes its value in a single 
grassmannian and we need to stratify $\mathcal W$ in order to get 
differentiable functions. For each integer $r \in [0,n]$, we introduce 
the sub-manifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
couples $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
induces a continuous function
$\II_r : \mathcal W_r \to \Grass(E,s)$.
Pick $(f,W) \in \mathcal W_r$. Set $V = f^{-1}(W)$ and denote by $w : F 
\to F/W$ the canonical projection.
Similarly to what we have done for direct images, one can prove that
the tangent space of $\mathcal W_r$ at some point $(f,W) \in \mathcal
W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
couples $(df,dw)$ such that
$d \tilde v = (w \circ df + dw \circ f)_{|W}$
factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
$f$. Furthermore $\II_r$ is differentiable at $(f,W)$ and its 
differential is the linear mapping that takes $(df,dw)$ as above to the 
unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
d\tilde v$.

@@

\subsection{Implementations and experiments}

@@

\noindent
{\bf Some experiments.}

@@

@@

Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
analytic around $0$ as well. Let us write the analytic expansion of $f$, 
$g$ and $h$ as follows:
$$f = \sum_{n \geq 0} f_n, \quad 
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n, \quad$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
aim of this subsection is to prove the following intermediate result.

@@

During the proof, multinomial coefficients will play an important role. 
We recall that they are defined as follows: 
given $s, k_1, \ldots, k_\ell$ some positive integers with $s \geq
k_1 + \cdots + k_\ell$, we set:
$$\binom s {k_1 \,\, k_2 \,\, \cdots \,\, k_\ell} =
  \frac{s!}{k_1!\: k_2! \cdots k_\ell! \: (s{-}k_1{-}\cdots{-}k_\ell)!}.$$
We also recall that these numbers are all positive integers.

We now expand $g \circ f$ as follows:
\begin{align}
g \circ f & = \sum_{m \geq 0} g_m \Big(\sum_{n \geq 1} f_n\Big) \nonumber \\
& = \sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
\label{eq:expansiongf}
\end{align}
where the latest sum runs over:

@@

We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
We write the analytic expansion of $f$, $g$ and $h$ as follows: 
$$f = \sum_{n \geq 0} f_n, \quad
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
Following \cite{caruso-roe-vaccon:14a}, we define $\NP(f) : \R \to \R \cup 
\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
$\Lambda(h)$ similarly.
We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
\Vert \geq e^{-\alpha n}$ for all positive integer $n$.

\begin{lem}
\label{lem:boundLambdaf}
We keep the above notations. If the couple $(a,b)$ satisfies:
\begin{equation}
\label{eq:condAB}
b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
\end{equation}
then $b \geq \Lambda(f)(a - \alpha)$.
\end{lem}

>>>>

\noindent
{\bf Organization of the paper.}
The section \ref{sec:theory} introduces the theory of ultrametric 
precision developped in \cite{caruso-roe-vaccon:14a} and completes 
it on a technical point.
In \S \ref{sec:matrices}, we focus on matrices... \todo{Complete
this}.

@@

In \cite{caruso-roe-vaccon:14a}, we also explained that if the function 
$f$ is locally analytic, then the constant $\delta$ appearing in
Proposition \ref{prop:precision} can be made explicit in terms of the
growing function $\Lambda(f)$ of $f$ defined by:
$$\textstyle \Lambda(f)(v) = 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert 
f(h) \Vert \big)$$
with the convention that $\Lambda(f)(v) = +\infty$ if $f$ does not
converge on $B^-_E(e^v)$.
We refer to \cite[Proposition 3.12]{caruso-roe-vaccon:14a} for the
precise statement and seek here to state the particular case of
integral polynomial functions. A function $f : E \to F$ is said 
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to orthonormal basis.

\begin{prop}
\label{prop:precision2}
We keep the notations of Proposition \ref{prop:precision} and assume 
in addition that $f$ is integral polynomial. Let $C$ be a positive real
number such that $B_F(1) \subset f'(v_0)(B_E(C))$. 
Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
\rho^{-1}$.
\end{prop}

In \cite[Appendix A]{caruso-roe-vaccon:14a}, the theory is extended 
to manifolds over $K$ and covers this way a very large families of
situations (\emph{e.g.} computation with vector spaces as discussed
in \S \ref{sec:vectorspaces}). In this extended framework, the precision 
datum at some point $x$ is a lattice in the tangent space at $x$. 
Propositions \ref{prop:precision} and \ref{prop:precision2} have
analogues as well, simply obtained by working in charts.

@@

We actually generalize a bit the setting above and consider the 
differential equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in two finite dimensional
normed vector spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a normed $K$-vector space $G$. Consequently,
$g$ goes from $V \times W$ to the vector space $\Hom(E,F)$ gathering
all (continuous) linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$ and second, in order to 
apply Proposition \ref{prop:precision2}, the derivative $f'(0)$ 
needs to be surjective and therefore \emph{a fortiori} nonzero.

@@

Actually, we do recommend taking $h$ as the identity function 
\emph{but} changing the norm on the codomain of $h$ at the same time.
More precisely, if $F$ is equipped with the norm $\Vert \cdot \Vert_F$, 
we recommand endowing it with the second norm defined by
$\Vert x \Vert'_F = \lambda \cdot \Vert x \Vert_F$ ($x \in F)$
where $\lambda$ is a positive real number that we shall choose later 
and taking $h : (F, \Vert \cdot \Vert_F) \to (F, \Vert \cdot \Vert'_F)$ 
acting as the identity on the underlying vector spaces. Doing this, the 
function $\Lambda(h)$ maps $x$ to $x + \log \lambda$ ($x \in \R$) and we 
choose $\Lambda_h = \Lambda(h)$.

@@

\subsection{Multiplication}
\label{subsec:mulmatrix}

To begin with, we want to study the behaviour of the precision when 
performing a matrix multiplication. Let $r$, $s$ and $t$ be three 
positive integers and assume that we want to multiply a matrix $A \in 
M_{r,s}(K)$ by a matrix $B \in M_{s,t}(K)$. This operation is of course 
modeled by the integral polynomial function:
$$\begin{array}{rcl}
\mathcal P_{r,s,t} : \quad M_{r,s}(K) \times M_{s,t}(K) & \to & 
M_{r,t}(K) \smallskip \\
(A,B) & \mapsto & AB.
\end{array}$$
According to Proposition \ref{prop:precision}, the behaviour of the precision when 
computed $AB$ is governed by $\mathcal P'_{r,s,t}(A,B)$. It is obvious 
that this differential is the linear mapping that takes a couple 
$(dA,dB)$ to $A \cdot dB + dA \cdot B$.

To fix ideas, let us assume from now that the entries of $A$ and $B$ all 
lie in $\O_K$ and are known at the same precision $O(\pi^N)$. In order 
to apply Propositions \ref{prop:precision} and \ref{prop:precision2}, we then need to compute the image 
of the standard lattice $\mathcal L_0 = M_{r,s}(\O_K) \times 
M_{s,t}(\O_K)$ under $\mathcal P'_{r,s,t}(A,B)$. It is of course 
contained in $M_{r,t}(\O_K)$; this reflects the obvious fact that each 
entry of the product $AB$ is also known with precision at least $O(\pi^N)$. 
Nonetheless, it may happen that the above inclusion is strict, meaning 
that we are \emph{gaining} precision in those cases. In order to compute 
exactly $\mathcal P'_{a,b,c}(A,B)(\mathcal L_0)$, we introduce the Smith 
factorization of $A$ and $B$:
$$A = U_A \cdot \Delta_A \cdot V_A
\text{ and } B = U_B \cdot \Delta_B \cdot V_B$$
and denote by $a_1, \ldots, a_{\min(r,s)}$ (resp. $b_1, \ldots, 
b_{\min(s,t)}$) the valuation of the diagonal entries of $\Delta_A$
(resp. $\Delta_B$). We also agree to set $a_n = b_m = +\infty$ if
$n > \min(r,s)$ and $m > \min(s,t)$. We define $M_{r,t}((a_i),(b_j))$ 
as the sublattice of $M_{r,t}(\O_K)$ consisting of matrices $M = (M_{i,j})$ 
such that $\val(M_{i,j}) \geq \min(a_i,b_j)$ for all $(i,j)$.

@@

Let us briefly comment on Proposition \ref{prop:mulmatrix} from the 
perspective of precision. The second statement roughly means that when 
we are computing the product $AB$, we are gaining $\sum_{m=1}^\infty 
a^{(m)} b^{(m)}$ significant digits in absolute precision\footnote{We 
note that, on the other side, the valuation of the entries of $AB$ may 
increase, meaning that we are also loosing some significant digits if we 
are reasoning in relative precision.} as soon as $N > \min(a_r, b_t)$ 
(\emph{cf} Proposition \ref{prop:precision2}). However these digits are 
not --- even individually --- attached to some particular entry of $AB$ 
but are ``spread out'' among all its entries. Thus, a usual 
coordinate-wise precision data does not see this gain of precision in 
general. There is a more pedestrian way to visualize this phenomenon. 
Remark that the product $AB$ can be rewritten:
$AB = U_A \cdot P \cdot V_B$ with
$P = \Delta_A \cdot V_A \cdot U_B \cdot \Delta_B$.
Tracking precision in the usual way, we see that the $(i,j)$-th entry
of $P$ is known at precision $O(\pi^{N + \min(a_i,b_j)})$. Hence, the 
gain of precision is visible here. However it is then ``spread out'' 
when we are computing $U_A \cdot P \cdot V_B$ and mostly disappears in 
the usual coordinate-wise precision model. The lattice precision model 
then appears as a way to keep track of this ``diffused'' gain of
precision.

@@

\noindent
Figure \ref{fig:mulmatrix} compares the number of significant digits in 
\emph{relative} precision we are loosing on the output of Algorithm 1 
when we are using, on the one hand, a standard coordinate-wise track of 
precision and, on the other hand, a lattice-based method to handle 
precision.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{2}{*}{\hspace{0.2cm}$d$\hspace{0.2cm}} & 
\multirow{2}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{2}{|c|}{Average loss of precision} \\
\cline{3-4}
& & Coord-wise method & Lattice method \\
\hline 
$2$ & $10$ & $\hphantom{00}2.8$ & $2.4$ \\
$2$ & $100$ & $\hphantom{0}16.7$ & $5.0$ \\
$2$ & $1000$ & $157.8$ & $7.9$ \\
\hline
$3$ & $10$ & $\hphantom{00}2.2$ & $1.9$ \\
$3$ & $100$ & $\hphantom{0}12.8$ & $4.0$ \\
$3$ & $1000$ & $122.5$ & $7.0$ \\
\hline
\end{tabular}

@@

\noindent
{\bf Direct images.}
Let $E$ and $F$ be two finite dimensional $K$-vector spaces of dimension 
$n$ and $m$ respectively. Let $d$ be an integer in $[0,n]$. We are
interested here in the function $\DI$ (for ``direct image'') defined 
over $\mathcal M = \Hom(E,F) \times \Grass(E,d)$ that takes the 
couple $(f,V)$ to $f(V)$. Since the dimension of $f(V)$ may \emph{a 
priori} vary, the map $\DI$ does not take its values in well-defined 
grassmannian. To tackle this problem, we stratify $\mathcal M$ as
follows: for each integer $r \in [0,d]$, we introduce the subset
$\mathcal M_r \subset \mathcal M$ consisting of those couples $(f,V)$
for which $f(V)$ has dimension $r$. The $\mathcal M_r$'s are
locally closed in $\mathcal M$ are therefore are submanifolds. 
Moreover, $\DI$ induces by restriction respectable differentiable 
functions between $K$-manifolds
$\DI_r : \mathcal M_r \to \Grass(F,r)$.

We would like to differentiate $\DI_r$ around some point $(f,V) \in 
\mathcal M_r$. To do so, we use the first description of the 
grassmannians we gave above: we see points in $\Grass(E,d)$ 
(resp. $\Grass(F,d)$) as embeddings $V \hookrightarrow E$ (resp. $W 
\hookrightarrow F$) modulo the action of $\GL(V)$ (resp. $\GL(W)$).
The point $V \in \Grass(E,d)$ is then represented by the canonical 
inclusion $v : V \to E$ whereas a representative $w$ of $W$ satisfies
$w \circ \varphi = f \circ v$
where $\varphi : V \to W$ is the linear mapping induced by $f$. The 
previous relation still holds if $(f,v)$ is replaced by a couple $(f', 
v') \in \mathcal M_r$ sufficiently close to $(f,v)$.
Differentiating it and passing to the quotient we find, first, that the 
tangent space of $\mathcal M_r$ at $(f,v)$ consists of couples $(df, dv) 
\in \Hom(E,F) \times \Hom(V,E/V)$ such that
$$d\tilde w = \big((df \circ v + f \circ dv) \text{ mod } W\big)
\in \Hom(V, F/W)$$
factors through $\varphi$ (\emph{i.e.} vanishes on $\ker \varphi = V 
\cap \ker f$) and, second, that the differential of $\DI_r$ at $(f,V)$ 
is the linear mapping sending $(df,dv)$ as above to the unique element 
$dw \in \Hom(W,F/W)$ such that $dw \circ \varphi = d \tilde w$.

@@

\noindent
{\bf Inverse images.}
We now consider the mapping $\II$ (for ``inverse image'') sending a 
couple $(f,W) \in \mathcal W = \Hom(E,F) \times \Grass(F,d)$ to 
$f^{-1}(W)$. As before, this map does not take its value in a single 
grassmannian and we need to stratify $\mathcal W$ in order to get 
differentiable functions. For each integer $r \in [0,n]$, we introduce 
the sub-manifold $\mathcal W_r$ of $\mathcal W$ consisting of those 
couples $(f,W)$ such that $\dim f^{-1}(W) = s$. For all $s$, $\II$ 
induces a continuous function
$\II_r : \mathcal W_r \to \Grass(E,s)$.
Pick $(f,W) \in \mathcal W_r$. Set $V = f^{-1}(W)$ and denote by $w : F 
\to F/W$ the canonical projection.
Similarly to what we have done for direct images, one can prove that
the tangent space of $\mathcal W_r$ at some point $(f,W) \in \mathcal
W_r$ is the subspace of $\Hom(E,F) \times \Hom(W,F/W)$ consisting of 
couples $(df,dw)$ such that
$d \tilde v = (w \circ df + dw \circ f)_{|W}$
factors through the linear mapping $\varphi : E/V \to F/W$ induced by 
$f$. Furthermore $\II_r$ is differentiable at $(f,W)$ and its 
differential is the linear mapping that takes $(df,dw)$ as above to the 
unique element $dv \in \Hom(V,E/V)$ satisfying $\varphi \circ dv =
d\tilde v$.

@@

\subsection{Implementation and experiments}

@@

\noindent
{\bf Some experiments.}
Let us consider the example of computation given by the following
algorithm.

\noindent {\bf Algorithm 2:} {\tt example\_vector\_space}

\noindent{\bf Input:} two integers $n$ and $N$

\noindent 1.\ Set $L_0 = \langle (1 + O(2^N),  O(2^N), O(2^N)) \rangle
\subset \Q_2^3$

\noindent 2.\ {\bf for} $i=0,\dots,n-1$

\noindent 3.\ \hspace{0.3cm}pick randomly $\alpha, \beta, \gamma, \delta
\in M_{3,3}(\Z_2)$ with high precision

\noindent 4.\ \hspace{0.3cm}{\bf compute} $L_{i+1} = 
\big(\alpha(L_i) + \beta(L_i)\big) \cap \big(\gamma(L_i) + \delta(L_i)\big)$

\noindent 5.\ {\bf return} $L_n$

\noindent
The expression ``high precision'' on line 3 means that the precision on 
$\alpha$, $\beta$, $\gamma$ and $\delta$ is set in such a way that it 
does not affect the resulting precision on $L_{i+1}$.
%
\begin{figure}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|c|c|c|c|}
\hline
\multirow{3}{*}{\hspace{0.2cm}$n$\hspace{0.2cm}} & 
\multicolumn{3}{|c|}{Average loss of precision} \\
\cline{2-4}
& \raisebox{-0.05cm}{Coord-wise} & 
\multicolumn{2}{|c|}{Lattice method} \\
\cline{3-4}
& \raisebox{0.04cm}{method} & 
\hspace{0.2cm}projected\hspace{0.2cm} & 
\hspace{0.2cm}diffused\hspace{0.2cm} \\
\hline 
$10$ & $\hphantom{0}7.3$ & $\hphantom{0}2.7$ & $\hphantom{0}{-}2.4 \times 2$ \\
$20$ & $14.8$ & $\hphantom{0}5.5$ & $\hphantom{0}{-}4.7 \times 2$ \\
$50$ & $38.6$ & $13.1$ & $-12.0 \times 2$ \\
$100$ & $78.1$ & $26.5$ & $-23.5 \times 2$ \\
\hline
\end{tabular}

{\small
Results for a sample of $1000$ executions (with $N \gg n$)}
\end{center}
\renewcommand{\arraystretch}{1}

\caption{Average loss of precision in Algorithm 2}
\label{fig:vectorspace}
\end{figure}
%
Figure \ref{fig:vectorspace} shows the losses of precision when
executing Algorithm 2 with various inputs $n$ (the input $N$ is
always chosen sufficiently large so that it does not affect the
behaviour of precision). The ``coord-wise column'' corresponds to
the standard way to track precision. On the other hand, in the two 
last columns, the precision is tracked using lattices. The meaning of 
``diffused'' is the same than those explained in \S 
\ref{subsec:mulmatrix} when we were dealing with matrix multiplication: 
it then corresponds to significant digits which are ``spread out'' among 
the two coordinates of the precision lattice. In the present situation,
it turns out that these ``diffused'' digits may be ``separated'' by
working in another basis of the underlying space $\Q_2^3$
The fact that only negative numbers appear in this column means that we 
are actually always gaining precision with this model! Finally, in the 
``projected'' column, we have written the loss of precision we get after 
having projected the lattice precision data onto each coordinate (in 
order to get an answer which is coordinate-wise).

@@

We prove Proposition \ref{prop:boundLambdaf} in the slightly more
general context of $K$-Banach.

@@

Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
analytic around $0$ as well. Let $f = \sum_{n \geq 0} f_n$
$g = \sum_{n \geq 0} g_n$ and $h = \sum_{n \geq 0} h_n$ be the
analytic expansions of $f$, $g$ and $h$.
Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
aim of this subsection is to prove the following intermediate result.

@@

A computation gives the following expansion for $g \circ f$:
\begin{equation}
\label{eq:expansiongf}
\sum \binom m {\!k_1 \,\, \cdots \,\, k_\ell\!} \:
G_m(f_{n_1}, \ldots, f_{n_1}, \ldots, f_{n_\ell}, \ldots, f_{n_\ell})
\end{equation}
where $\binom m {\!k_1 \,\, \cdots \,\, k_\ell\!}$ denotes the
multinomial coefficient and the sum runs over:

@@

We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
Let $f = \sum_{n \geq 0} f_n$, $g = \sum_{n \geq 0} g_n$ and
$h = \sum_{n \geq 0} h_n$ be the analytice expansions of
$f$, $g$ and $h$.
Here $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
Following \cite{caruso-roe-vaccon:14a}, we define $\NP(f) : \R \to \R \cup 
\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
\log \Vert f_n \Vert$ for all integer $n$. We recall that $\Lambda(f)$ 
is the Legendre transform of $\NP(f)$ 
\cite[Proposition 3.9]{caruso-roe-vaccon:14a}.
We define $\NP(g)$ and $\NP(h)$ similarly.
We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
\Vert \geq e^{-\alpha n}$ for all positive integer $n$.

\begin{lem}
\label{lem:boundLambdaf}
We keep the above notations. 
If $(a,b)$ satisfies
$b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)$
then $b \geq \Lambda(f)(a - \alpha)$.
\end{lem}
